---
title: "EDA"
author: "Tory Norton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(patchwork)
library(RColorBrewer)
#load data
load("particleAdhesion.RData")
adhesion_original <- x
theme_set(theme_classic())
```


# data wrangling

```{r}
# remove constants and redundant cols
adhesion_small <- adhesion_original %>%
  select(-c(bins, removed, hamaker, density, workOfAdhesion))

# move "original" counts to rows with RPM 0
adhesion <- adhesion_small %>%
  group_by(site, sample, avgDiameter, type) %>%
  summarise(retained = unique(original), .groups = "drop") %>%
  mutate(rotation = 0) %>%
  rbind(select(adhesion_small, -original))

# calculate area for each sample/site
adhesion_area <- adhesion %>%
  # total diameter/area for each size
  mutate(diam = avgDiameter * retained,
         area = (avgDiameter) ^ 2 * pi / 4 * retained) %>%
  group_by(site, sample, rotation, type) %>%
  summarise(
    # sum over sizes
    total_diam = sum(diam),
    total_area = sum(area),
    retained = sum(retained),
    .groups = "drop"
  ) %>%
  mutate(
    # calculate avg diam & total area
    avg_diam = case_when(retained == 0 ~ 0, TRUE ~ total_diam / retained),
    avg_area = case_when(retained == 0 ~ 0, TRUE ~ total_area / retained),
    unique_id = paste0(type, sample, site),
    total_area_cm = total_area/1e+4,
    avg_area_cm = avg_area/1e+4
  )
```


## choosing response variable


```{r}
value_labeller <- c(
  "avg_diam" = "'Avg. diameter ('*mu*'m)'",
  "retained" = "'Count'",
  "total_area_cm" = "'Total area (cm'^2*')'"
)
particle_types <- adhesion_area$type %>% unique()

plot_count_diam_area <- function(ptype, legend = F) {
  value_labeller <- c(
    "avg_diam" = "'Avg. diameter ('*mu*'m)'",
    "retained" = "'Count'",
    "total_area_cm" = "'Total area (cm'^2*')'"
  )
  particle_types <- adhesion_area$type %>% unique()
  if (!ptype %in% particle_types) {stop("Invalid type")}
  p <- adhesion_area %>%
    filter(type == ptype) %>%
    mutate(avg_diam = case_match(avg_diam, 0 ~ NA, .default = avg_diam)) %>%
    pivot_longer(cols = c("retained", "avg_diam", "total_area_cm")) %>%
    mutate(name = fct_relevel(name, "retained")) %>%
    ggplot(aes(
      x = rotation,
      y = value,
      color = as.factor(sample)
    )) +
    geom_smooth(se = F, linewidth = 0.7) +
    geom_jitter() +
    facet_wrap(
      ~ name,
      labeller = as_labeller(value_labeller, default = label_parsed),
      ncol = 1,
      scales = "free_y"
    ) +
    scale_color_brewer(type = "qual", palette = 3) +
    labs(title = ptype, color = "Sample", # y = "",
         x = "RPM") +
    theme_classic() +
    theme(
      panel.grid.major = element_line(),
      axis.title.y = element_blank()
    )
  if (!legend) {
    p <- p +
      theme(legend.position = "none")
  }
  return(p)
}

p1 <- plot_count_diam_area("alfalfa")
p2 <- plot_count_diam_area("gypsum", legend = T)

(p1 | p2)
```



```{r, message=FALSE, warning = F}
rpm <- adhesion_area$rotation %>% 
  unique()

fig_area_by_type <- adhesion_area %>%
  # remove outlier for plotting
  filter(type != "flyash" | sample != 3 | rotation != 0 | site != 4) %>%
  ggplot(aes(
    x = rotation,
    y = total_area_cm,
    color = as.factor(sample)
  )) +
  geom_smooth(se = F, linewidth = 0.6) +
  geom_jitter() +
  facet_wrap( ~ type, scales = "free_y") +
  scale_x_continuous(
    breaks = rpm, 
    labels = rpm/1000
  ) +
  theme_classic() +
  labs(
    color = "Sample",
    x = "RPM (thousands)",
    y = expression("Area (cm"^2*")"),
    title = "Particle coverage area after each round of centrifuging"
    ) +
  theme(legend.position = "right") +
  scale_color_brewer(type = "qual", palette = 3)
fig_area_by_type
```







# cleaning: identifying outliers

## High outlier in flyash

```{r}
# crazy outlier
outlier_data <- adhesion_area %>%
  filter(type == "flyash" & sample == 3 & rotation == 0 & site == 4)

adhesion_area %>%
  # remove outlier for plotting
  filter(type != "flyash" | sample != 3 | rotation != 0 | site != 4) %>%
  ggplot(aes(
    x = rotation,
    y = total_area_cm,
    color = as.factor(sample)
  )) +
  geom_smooth(se = F, linewidth = 0.6) +
  geom_jitter() +
  facet_wrap(~ type) +
  scale_x_continuous(breaks = rpm, labels = rpm / 1000) +
  theme_classic() +
  labs(
    color = "Sample",
    x = "RPM (thousands)",
    y = expression("Area (cm" ^ 2 * ")"),
    title = "Particle coverage area after each round of centrifuging"
  ) +
  theme(legend.position = "right") +
  scale_color_brewer(type = "qual", palette = 3) +
  geom_point(data = outlier_data) +
  geom_point(data = outlier_data, shape = 1, color = "red", size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.2)))

```


```{r}
p_graphite <- adhesion_area %>%
  # remove outlier for plotting
  filter(type == "graphite") %>%
  ggplot(aes(
    x = rotation,
    y = total_area_cm,
    color = as.factor(sample),
    group = unique_id
  )) +
  geom_line() +
  geom_jitter() +
  scale_x_continuous(breaks = rpm, labels = rpm / 1000) +
  theme_classic() +
  labs(
    color = "Sample",
    x = "RPM (thousands)",
    y = expression("Area (cm" ^ 2 * ")"),
    title = "graphite"
  ) +
  scale_color_brewer(type = "qual", palette = 3) 

p_graphite1 <- adhesion_area %>%
  filter(type == "graphite" & sample == 1) %>% 
  ggplot(aes(
    x = rotation,
    y = total_area_cm,
    color = as.factor(sample),
    group = unique_id
  )) +
  facet_wrap(~site)+
  geom_line() +
  geom_jitter() +
  scale_x_continuous(breaks = rpm, labels = rpm / 1000) +
  theme_classic() +
  labs(
    color = "Sample",
    x = "RPM (thousands)",
    y = expression("Area (cm" ^ 2 * ")"),
    title = "Sample 1 by site"
  ) +
  scale_color_brewer(type = "qual", palette = 3) 
  
p_graphite 
p_graphite1

adhesion %>%
  filter(type == "graphite" & sample == 1 & rotation %in% c(0, 10000)) %>%
  uncount(retained, .remove = F) %>% 
  ggplot(aes(
    x = avgDiameter,
    fill = as.factor(rotation),
    color = as.factor(rotation)
    # alpha = as.factor(rotation)
  )) +
  geom_histogram(
    position = "identity", 
    # color = "gray15", 
    linewidth = 0.2,
    binwidth = 0.3
    ) +
  facet_wrap(
    ~site, 
    scales = "fixed",
    labeller = as_labeller(\(x) paste("Site", x))
    ) +
  scale_fill_manual(
    values = c(alpha("steelblue3", 0.7), alpha("firebrick", 0.5)),
    aesthetics = c("fill", "color"),
    labels = c("Original", "10000rpm")
    ) +
  scale_x_continuous(transform = "sqrt") +
  labs(title = "graphite sample 1",
       # y = "Site",
       x = expression(sqrt("diameter"))) +
  theme(legend.title = element_blank(),
        legend.position = c(1,1),
        legend.justification = c(1,1),
        legend.key.size = unit(0.03, "snpc"),
        legend.key.spacing.y = unit(0.01, "npc"),
        )

# adhesion_original %>%
#   filter(type == "graphite" & sample == 1 & rotation == 10000 & site == 3) %>% 
#   ggplot(aes(x = avgDiameter, y = removed)) +
#   geom_hline(yintercept = 0) +
#   geom_point()


adhesion_area %>%
  filter(type == "LQ") %>%
  ggplot(aes(
    x = rotation,
    y = total_area_cm,
    color = as.factor(sample),
    group = unique_id
  )) +
  geom_line() +
  geom_jitter() +
  scale_x_continuous(breaks = rpm, labels = rpm / 1000) +
  theme_classic() +
  labs(
    color = "Sample",
    x = "RPM (thousands)",
    y = expression("Area (cm" ^ 2 * ")"),
    title = "LQ"
  ) +
  scale_color_brewer(type = "qual", palette = 3) 

adhesion_original %>%
  filter(type == "LQ" & sample == 5 & site == 1 & rotation == 1000) %>% 
  pivot_longer(c(retained, original)) %>% 
  ggplot(aes(x = avgDiameter, y = value, fill = name)) +
  # geom_hline(yintercept = 0) +
  geom_col(position = "dodge") +
  # geom_point() +
  theme_bw() +
  scale_x_continuous(limits = c(0,50)) +
  labs(title = "LQ sample 5 site 1")
```


- graphite sample 1, site 3 has unusual area values for 0 and 10000rpm
  - when we look at counts for each particle size, nearly all of them are larger after 10000rpm compared to original counts
  - doesn't make sense -- counts appear to have been measured or entered incorrectly
  - we will remove these measurements

- removing "original" measurement for LQ sample 5 site 1 -- does not make sense for count of largest particle size to increase drastically after spinning at 1000rpm


```{r}
outlier_ids <- tribble(
  ~unique_id, ~rotation,
  "flyash34", 0,
  "LQ51", 0,
  "graphite13", 0,
  "graphite13", 10000
)

adhesion_outliers <- adhesion_area %>%
  semi_join(outlier_ids)
adhesion_clean <- adhesion_area %>%
  anti_join(outlier_ids)

adhesion_clean %>%
  filter(type %in% adhesion_outliers$type) %>% 
  ggplot(aes(
    x = rotation,
    y = total_area_cm,
    color = as.factor(sample)
  )) +
  geom_smooth(se = F, linewidth = 0.6) +
  geom_jitter() +
  facet_wrap(~ type, scales = "free") +
  scale_x_continuous(breaks = rpm, labels = rpm / 1000) +
  theme_classic() +
  labs(
    color = "Sample",
    x = "RPM (thousands)",
    y = expression("Area (cm" ^ 2 * ")"),
    title = "Particle coverage area after each round of centrifuging"
  ) +
  theme(legend.position = "right") +
  scale_color_brewer(type = "qual", palette = 3) +
  geom_point(data = adhesion_outliers) +
  geom_point(data = adhesion_outliers, shape = 1, color = "red", size = 6, stroke = 1)
  # scale_y_continuous(transform = "log")
```
Four outliers removed


# LME Model

```{r}
adhesion_clean <- adhesion_clean %>% 
  mutate(area_cm = total_area_cm,
         area_cm_log = log1p(area_cm),
         area_cm_sqrt = sqrt(area_cm))

adhesion_clean %>% 
  filter(!type %in% c("flyash", "SiO2", "graphite")) %>% 
  pivot_longer(
    starts_with("area"),
    names_to = "transform"
    ) %>% 
  ggplot(aes(
    x = rotation,
    y = value,
    color = as.factor(sample)
  )) +
  geom_smooth(se = F, linewidth = 0.5) +
  geom_jitter(alpha = 0.5) +
  # facet_wrap( ~ type, scales = "fixed") +
  facet_grid(rows = vars(transform), cols = vars(type), scales = "free_y") +
  scale_x_continuous(
    breaks = rpm, 
    labels = rpm/1000
  ) +
  scale_color_brewer(type = "qual", palette = 3) +
  theme_bw() +
  theme(legend.position = "top")
```

- considered root and log-transform for area
- goal: "linearity in RPM"
  - both transforms help a bit for 0 - 1000rpm, but make the graphs look more nonlinear when counts drop to 0

## physical constants

```{r}
adhesion_constants <- adhesion_original %>% 
  group_by(type) %>% 
  summarise(
    density = unique(density),
    hamaker = unique(hamaker),
    workOfAdhesion = unique(workOfAdhesion)
    )

adhesion_constants %>% 
  ggplot(aes(x=hamaker, y = workOfAdhesion, color = density)) + 
  geom_point()

adhesion_clean <- adhesion_clean %>% 
  left_join(adhesion_constants)
```

- hamaker and WOA are completely correlated, so we consider only density and WOA


```{r}
adhesion_clean %>% 
  filter(rotation == 0) %>% 
  ggplot(aes(y = total_area_cm, x = density, color = type, group = 1)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Density")

adhesion_clean %>% 
  filter(rotation == 0) %>% 
  ggplot(aes(y = total_area_cm, x = workOfAdhesion, color = type, group = 1)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "workOfAdhesion")
  
```

Little evidence that density has a linear effect on intercept (area at 0 rpm), but some evidence for workOfAdhesion


```{r}
adhesion_clean %>% 
  # filter(!type %in% c("gypsum", "solosphere")) %>%
  pivot_wider(
    id_cols = c(unique_id, site, sample, type), 
    names_from = rotation, 
    values_from = total_area_cm,
    names_prefix = "rpm_") %>% 
  left_join(adhesion_constants) %>% 
  mutate(area_rm_1000 = rpm_0 - rpm_1000) %>% 
  pivot_longer(cols = c(density, workOfAdhesion)) %>% 
  ggplot(aes(x = value, y = area_rm_1000, color = type, group = name)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "all types")

adhesion_clean %>% 
  filter(!type %in% c("gypsum")) %>%
  pivot_wider(
    id_cols = c(unique_id, site, sample, type), 
    names_from = rotation, 
    values_from = total_area_cm,
    names_prefix = "rpm_") %>% 
  left_join(adhesion_constants) %>% 
  mutate(area_rm_1000 = rpm_0 - rpm_1000) %>% 
  pivot_longer(cols = c(density, workOfAdhesion)) %>% 
  ggplot(aes(x = value, y = area_rm_1000, color = type, group = name)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = F) +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "without gypsum")

adhesion_clean %>% 
  filter(!type %in% c("gypsum", "solosphere")) %>%
  pivot_wider(
    id_cols = c(unique_id, site, sample, type), 
    names_from = rotation, 
    values_from = total_area_cm,
    names_prefix = "rpm_") %>% 
  left_join(adhesion_constants) %>% 
  mutate(area_rm_1000 = rpm_0 - rpm_1000) %>% 
  pivot_longer(cols = c(density, workOfAdhesion)) %>% 
  ggplot(aes(x = value, y = area_rm_1000, color = type, group = name)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "without gypsum and solosphere")
```

- some evidence that both density and workOfAdhesion affect the change in area after centrifuging
- gypsum and solosphere have unusually high values and high variance
- however, the linear effect of WOA appears much weaker when gypsum is removed, and changes direction when gypsum and solosphere are removed
  - linear effect of density is always positive

# Model


```{r}
library(lme4)
diam_rpm0 <- filter(adhesion_area, rotation == 0) %>% 
  group_by(type) %>% 
  summarise(total_diam = sum(total_diam), retained = sum(retained), avg_diam0 = total_diam/retained)

adhesion_clean <- adhesion_clean %>% 
  mutate(density_scaled = scale(density), 
         woa_scaled = scale(workOfAdhesion),
         samp_id = paste0(type,sample)
         ) %>% 
  left_join(select(diam_rpm0, type, avg_diam0))

mod_lm <- lm(
  total_area_cm ~ woa_scaled + density_scaled*rotation + rotation,
  data = adhesion_clean
  )


mod_type_int <- lmer(
  total_area_cm ~ (1|type) + woa_scaled + rotation + density_scaled*rotation,
  data = adhesion_clean
  )

mod_type_slope0 <- lmer(
  total_area_cm ~ (1 + rotation|type) + woa_scaled + rotation +  density_scaled*rotation,
  data = adhesion_clean
  )
mod_type_slope1 <- lmer(
  total_area_cm ~ (1 + rotation|type) + woa_scaled + rotation + avg_diam0*rotation +  density_scaled*rotation,
  data = adhesion_clean
  )

mod_site_int <- lmer(
  total_area_cm ~ (1|unique_id) + woa_scaled + rotation + density_scaled*rotation,
  data = adhesion_clean
  )

mod_samp_slope <- lmer(
  total_area_cm ~ (1|samp_id) + (rotation|type) + woa_scaled + rotation + density_scaled*rotation,
  data = adhesion_clean
  )

mod_site_slope <- lmer(
  total_area_cm ~ (1|unique_id) + (rotation|type) + woa_scaled + rotation + density_scaled*rotation,
  data = adhesion_clean
  )

mod_list <- list(
  mod_lm,
  mod_type_int,
  mod_type_slope0,
  mod_type_slope1,
  mod_site_int,
  mod_samp_slope,
  mod_site_slope
)

tibble(
  model = c("mod_lm", "mod_type_int", "mod_type_slope0", "mod_type_slope1", "mod_site_int", "mod_samp_slope", "mod_site_slope"),
  AIC = sapply(mod_list, AIC),
  BIC = sapply(mod_list, BIC)
)

plot(mod_site_slope)

re_df_site <- broom.mixed::tidy(mod_site_slope, effects = "ran_vals")

library(lattice)
dotplot(ranef(mod_site_slope, condVar = TRUE),
        strip = FALSE)

ggplot(re_df_site, aes(x = estimate, y = level)) +
  geom_point() +
  facet_wrap(~ term, scales = "free") +
  labs(x = "Random effect estimate",
       y = "Group level") +
  theme_bw()

re_df_type <- broom.mixed::tidy(mod_type_slope, effects = "ran_vals")

ggplot(re_df_type, aes(x = estimate, y = level)) +
  geom_point() +
  facet_wrap(~ term, scales = "free") +
  labs(x = "Random effect estimate",
       y = "Group level") +
  theme_bw()

coef(mod_type_int)
coef(mod_type_slope)
```

```{r}
mod_site_slope <- lmer(
  total_area_cm ~ (1|unique_id) + (rotation|type) + woa_scaled + rotation + density_scaled*rotation,
  data = adhesion_clean
  )
```


All predictors/response scaled 

$p$ - particle type
$i$ - sample
$j$ - rotation

$$
\begin{aligned}
y_{pij} &= \beta_0+ \beta_1\text{WOA}_p + \beta_2\text{rpm}_j + \beta_3(\text{rpm}_j\times \text{density}_p) + \mu_{i} + \gamma_i \text{rpm}_j + \epsilon_j \\
\beta_k &\sim N(0, \sigma_\beta^2) \\
\mu_i &\sim N(0, \sigma_\mu^2) \\
\gamma_i &\sim N(0, \sigma_\gamma^2) \\
\sigma_\ell &\sim \text{half-t}(0, 0.01) \qquad \ell \in \{\beta, \mu,\gamma\}\\
\epsilon_j &\sim N(0, \sigma^2) \\
\sigma &\sim \text{half-t}(0, 0.01)
\end{aligned}
$$



